Vertical and oblique transmission with fluctuating transmission

### 2018-06-03 ms/ms_sunnyvale.tex
finished real roots; todos and revisions discussed with MWF

@@

For the simple case of symmetric migration $m_1=m_2=m<1/2$ after symmetric selection $w_1=1/w_2=w>1$, the recursions~\eqref{eq:migration_model} become
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1'' &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2'' &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

@@

For the stability of the equilibria when $\rho_1=\rho_2=1$, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $f(\lambda) = a \lambda^2 + b \lambda + c$ with
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w^2+1)}{w}, \quad
c &=  1-2m.
\end{aligned}
\end{equation}
First, the discriminant $\Delta = b^2-4ac$ is positive for any real $m$, because, when writing it as a quadratic polynomial in $m$, $\Delta(m)=\frac{1}{w^{2}} \big(m^{2} (w^2+1)^2 - 2 m (w^2-1)^2 + (w^2-1)^2\big)$, this polynomial is positive at $m=0$ and $m=1/2$ and has a negative discriminant $-16w^2(w^2-1)^2$.
Therefore, the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two positive roots. \\
Second, as $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is greater than one, both equilibria are unstable, and a protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_polymorphism}.

When $\rho_1=1$ and $\rho_2=\rho$, the stability of $(0,0)$ (i.e. fixation in $B$) is determined by the characteristic polynomial $f(\lambda)$ with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m) (w + 1-\rho + \rho/w), \quad
c &=  (1-2m)(w(1-\rho)+\rho).
\end{aligned}
\end{equation}

First, the discriminant is 
$$
\Delta = (1 - m)^2 (w + 1-\rho + \rho/w)^2 - 4(1-2m)(w(1-\rho)+\rho),
$$
which is positive for $m=0$ and $m=1/2$:
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w + \rho) (w-1)}{w}\Big)^2 \\
\Delta(m=1/2) = \Big(\frac{w^2 + w(1-\rho) + \rho}{2 w}\Big)^2.
\end{aligned}\end{equation}
In addition, if we write $\Delta$ as a polynomial in $m$, its discriminant is negative -- $- 16 (w + \rho)^2 (w - 1)^2 \big(w (1 - \rho) + \rho \big) / w^2$, and therefore the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$ and stability can be analyzed by the sign of $a+b+c$.
For $\rho=1$ we already have $a+b+c<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any valid $\rho$, the leading eigenvalue is greater than one, and $(0,0)$ is unstable.

The stability of $(1,1)$ (i.e. fixation in $A$) is determined by the characteristic polynomial with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1-m)(w \rho + 1 -\rho + 1/w), \quad
c &= (1-2m) \big((1-\rho)/w + \rho\big),
\end{aligned}
\end{equation}
%[MWF: Are the roots real?]
Again, $a>0$, $b<0$, and $c>0$, and if $\rho=1$ then $a+b+c<0$, as before.
However, for $\rho=0$ we have $a+b+c = \frac{m}{w}(w-1)>0$, in which case the stability is determined by $2a+b=\frac{m(w+1)+w-1}{w}$, the derivative of the polynomial at $\lambda=1$, which is positive, and therefore both eigenvalues occur in $(0,1)$ and fixation is stable. %

@@

% Facebook rural internet 
% National service - Israel, Singapore
% Immigrtion - draught
% Opening schools, universities
% sending students abroad, bringing experts from abroad
% foreign labor
% refugees
% hunter-gatherers/chimps increase habitat in years with less food production
% religion - dark ages etc.

>>>>

% TODO what can of spatial differences can we imagine in transmission?
% two populations, one has first cousin marriages and one doesn't (endogmia)
% orthodox religious communities
% boarding schools for one population
% foraging in groups or families
% pigmies women 'migration' - adoption by husband family
% interesting: migration correlated with transmission rate, future work

% matrilineal primates - vertical transmission stronger in females than males
% similar differences in ultra religious human societies

@@

For the simple case of symmetric migration $m_1=m_2=m<1/2$ occurring after symmetric selection $w_1=1/w_2=w>1$, the recursions~\eqref{eq:migration_model} become
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1'' &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2'' &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

@@

For the stability of the equilibria when $\rho_1=\rho_2=1$, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $f(\lambda) = a \lambda^2 + b \lambda + c$ with
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w^2+1)}{w}, \quad
c &=  1-2m.
\end{aligned}
\end{equation}
First, the discriminant $\Delta = b^2-4ac$ is positive for any real $m$, because, when writing it as a quadratic polynomial in $m$, $\Delta(m)=\frac{1}{w^{2}} \big(m^{2} (w^2+1)^2 - 2 m (w^2-1)^2 + (w^2-1)^2\big)$, this polynomial is positive at $m=0$ and $m=1/2$ and has a negative discriminant $\big(-16w^2(w^2-1)^2\big)$.
Therefore, the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two positive roots. \\
Second, as $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is greater than one, both equilibria are unstable, and a protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_polymorphism}.

When $\rho_1=1$ and $\rho_2=\rho$, the stability of $(0,0)$ (i.e. fixation in $B$) is determined by the characteristic polynomial $f(\lambda)$ with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m) (w + 1 - \rho + \rho/w), \quad
c &=  (1 - 2m)(w(1-\rho)+\rho).
\end{aligned}
\end{equation}

First, the discriminant is 
$$
\Delta = (1 - m)^2 (w + 1-\rho + \rho/w)^2 - 4(1-2m)(w(1-\rho)+\rho),
$$
which is positive for $m=0$ and $m=1/2$:
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w + \rho) (w-1)}{w}\Big)^2 \\
\Delta(m=1/2) = \Big(\frac{w^2 + w(1-\rho) + \rho}{2 w}\Big)^2.
\end{aligned}\end{equation}
In addition, if we write $\Delta$ as a polynomial in $m$, its discriminant is negative $\big(- 16 (w + \rho)^2 (w - 1)^2 \big(w (1 - \rho) + \rho \big) / w^2\big)$, and therefore the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$ and stability can be analyzed by the sign of $a+b+c$.
For $\rho=1$ we already have $a+b+c<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any valid $\rho$, the leading eigenvalue is greater than one, and $(0,0)$ is unstable.

The stability of $(1,1)$ (i.e. fixation in $A$) is determined by the characteristic polynomial $f(\lambda)$ with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m)(w \rho + 1 - \rho + 1/w), \quad
c &= (1 - 2m) \big((1-\rho)/w + \rho\big),
\end{aligned}
\end{equation}
First, the discriminant is 
$$
\Delta = (1 - m)^2(w \rho + 1 - \rho + 1/w)^2 - 4 (1 - 2m) \big((1-\rho)/w + \rho\big),
$$
which is positive for $m=0$ and $m=1/2$:
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w \rho + 1) (w-1)}{w}\Big)^2 \\
\Delta(m=1/2) = \Big(\frac{w^2 \rho + w(1-\rho) + 1}{2 w}\Big)^2.
\end{aligned}\end{equation}
In addition, if we write $\Delta$ as a polynomial in $m$, its discriminant is negative $\big(- 16 (w\rho + 1)^2 (w - 1)^2 \big(\rho w + 1 - \rho \big) / w^3\big)$, and therefore the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$, and if $\rho=1$ then $a+b+c<0$, as before.
However, for $\rho=0$ we have $a+b+c = \frac{m}{w}(w-1)>0$, in which case the stability is determined by $2a+b=\frac{m(w+1)+w-1}{w}$, the derivative of the polynomial at $\lambda=1$, which is positive, and therefore both eigenvalues occur in $(0,1)$ and fixation is stable. %

@@

% Facebook rural internet 
% National service - Israel, Singapore
% Boarding schools
% Immigrtion - draught
% Opening schools, universities
% sending students abroad, bringing experts from abroad
% foreign labor
% refugees
% hunter-gatherers/chimps increase habitat in years with less food production
% religion - dark ages etc.
